---
title: "U.S. Weather Events Impact Analysis"
author: "DanGuti"
date: "February 7, 2016"
output: html_document
---
**Synopsis**  
The analysis is started by retrieving weather event data and slimming down the 
data to only necessary variables.

#Data Processing
First step for the analysis is to download the data and assign it to a usable 
environment variable.
```{r, cache=TRUE}
u <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2" #file url address
download.file(u,"StormData.csv.bz2") #download zip file
sData <- read.csv("StormData.csv.bz2") #read csv data

str(sData) #get some idea what variables are available.
require(plyr)
require(ggplot2)
require(reshape2)
```

## Data Reduction
Now we want to make things easier for our machines by getting rid of the 
variables that are not of interest to us.

```{r}
intvars <- c("EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP", 
             "CROPDMG", "CROPDMGEXP") #index of variables of interest
sData <- sData[intvars] #reduce data to only relevant variables
str(sData) #to show our reduced variables
```
## Data Cleaning
It seems like the crop damage exponential and property damage exponentials have 
some invalid values (?, -, +).  
Here it is assumed that the total value of property damages is of the type 
PROPDMG * 10 ^ PROPDMGEXP.
Under this assumption it would be of our interest to make this variable entirely 
numeric, so that k = 3, h=2 and m=9 and so on.  
Let's clean the exponential variables so that they are more practical to use.

```{r}
require(plyr)
require(ggplot2)
require(reshape2)
cData <- data.frame("EVTYPE" = sData$EVTYPE, "FATALITIES" = sData$FATALITIES, 
                    "INJURIES" = sData$INJURIES)## Initialize cData frame where clean exponential values will be entered

## Making the damage exponentials uniform numerical data
### Replace levels for property damage exponential
levels(sData$PROPDMG) #look at the levels we have from property damage exponentials

replacing <- c("?" = NA, 
              "-" = 0, "+" = 0, 
              "h" = 2, "H" = 2, 
              "K" = 3,
              "m" = 6, "M" = 6,
              "B" = 9) #list of value equivalencies to transform letters to number exponential
cleanPDMGEXP <- revalue(sData$PROPDMGEXP, replacing) #transform letters to exponential values. The given number is assumed to be of the form 10^x.

lv <- levels(cleanPDMGEXP) #save levels of property damage exponential
lv[1] = NA #when no damage is reported, list as NA
levels(cleanPDMGEXP) = lv #Fix levels on cleanPDMGEXP
cleanPDMGEXP = as.numeric(levels(cleanPDMGEXP)[cleanPDMGEXP]) #converts factor variable to numeric
cData$totPdmg <- sData$PROPDMG * 10^cleanPDMGEXP
```
So far, we cleaned the property damage exponential variable of non-valid values,
then changed all values to numeric; and finally, created a 
total property damage variable that gives a dollar amount for property damage.  
Our next step is to do the same for the corresponding crop damage variables.

```{r}
### Replace levels for crop damage exponential
levels(sData$CROPDMGEXP) #look at the levels we have from property damage exponentials

replacing <- c("?" = NA, 
               "k" = 3, "K" = 3,
               "m" = 6, "M" = 6,
               "B" = 9) #list of value equivalencies to transform letters to number exponential
cleanCDMGEXP <- revalue(sData$CROPDMGEXP, replacing) #transform letters to exponential values. The given number is assumed to be of the form 10^x.

lv <- levels(cleanCDMGEXP) #save levels of property damage exponential
lv[1] = NA #when no damage is reported, list as NA
levels(cleanCDMGEXP) = lv #Fix levels on cleanPDMGEXP
cleanCDMGEXP = as.numeric(levels(cleanCDMGEXP)[cleanCDMGEXP])#converts factor variable to numeric, the -1 offsets the value to match the levels
cData$totCdmg <- sData$CROPDMG * 10^cleanCDMGEXP

### Finally, we change NA's to 0 in the totCdmg and totPdmg variables, so that we only have dollar amounts and can come up with one single damage No.
cData$totPdmg[is.na(cData$totPdmg)] <- 0 #change NA's to 0
cData$totCdmg[is.na(cData$totCdmg)] <- 0 #change NA's to 0
cData$totDmg <- cData$totPdmg + cData$totCdmg #create totals column for damage
```
Now we have clean and useful variables gathered in cData. Let's take a look:

```{r}
head(cData)
```

This looks much better now. Let's see what we can find out about weather events.

#Data Summary

Now we can produce a useful summary of the data that we are interested in.
We aggregate data by event type, so that we know how many fatalities and injuries
each specific type of event causes.
```{r}
## Harmful events - Injuries or Fatalities
Harm <- aggregate(cData[,2:3], list(EVTYPE = cData$EVTYPE), sum) #create table with fatality and Injury aggregates


## Damage
Dam <- aggregate(cData[,4:6], list(EVTYPE = cData$EVTYPE), sum) #aggregate fatalities

```
With all the data available to us, let's take a look at the results.

# Results
Now let's order the events my most harmful, and most damaging.
Let's consider most harmful to be those with most fatalities, and similarly
most damaging to be those with most *total* damages in $ amount.

```{r}
TopHarm <- head(arrange(Harm,desc(Harm$FATALITIES))) #retrieve 6 EVTYPEs with most fatalities
TopDam <- head(arrange(Dam,desc(Dam$totDmg))) #retrieve 6 EVTYPEs with most total damage
TopHarm
TopDam
```
Now we can observe the most damaging and harmful events as shown in the tables above.
Finally, let's plot that data for a more visual representation of the data.

```{r}
TH <- melt(TopHarm, id = "EVTYPE")


ggplot(data=TH, aes(x=EVTYPE, y=value, fill=variable)) + 
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle=60,hjust=0.9,vjust=0.9))+
  ggtitle("Weather Event Harm") +
  labs(x="Event Type",y="Fatalities or Injuries") 
```
  
  Figure 1. Number of fatalities/injuries by event type.  
    
  
  
```{r}
TD <- melt(TopDam[,1:3], id = "EVTYPE")
ggplot(data=TD, aes(x=EVTYPE, y=value, fill=variable)) + 
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle=60,hjust=0.9,vjust=0.9))+
  ggtitle("Weather Event Damages") +
  labs(x="Event Type",y="Value of Damages ($)") 
```
  
  Figure 2. Damage in USD per event type.  
  
  
From these plots it is easier to conclude a few things:

1. The most harmful event is no doubt tornadoes, with the most fatalities and injuries.
2. The most damaging event on the other hand are floods, with almost double the amount of total damages as the second place.
3. Within the events that cause the most damage, property damage is by far the most significant contribution to the damages total.
4. Tornadoes with the most casualties by far, and a third place in damages is quite possibly the most relevant type of event.

